\huge \textbf{Words That Matter:\\ Vocabulary Pruning in ModernBERT}

### 2025-05-14 paper/main.tex
paper sections methodology and experiments

% !TEX root = main.tex
\documentclass[twocolumn]{article}
\usepackage[a4paper,margin=1in]{geometry} 
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{mathptmx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{arydshln}
\usepackage{float}
\usepackage{cite}
\usepackage{multirow}

@@

\begin{abstract}
Large language models like ModernBERT face deployment challenges due to their computational demands and memory footprint, limiting their use in resource-constrained environments. While various pruning techniques exist, most require resource-intensive fine-tuning, creating a bottleneck for efficient model adaptation. We introduce a novel vocabulary pruning approach for ModernBERT that requires no additional fine-tuning, delivering superior efficiency particularly at smaller pruning ratios. Our technique specifically targets the embedding layer, which constitutes approximately 25\% of the model's parameters, through a systematic token importance analysis. We demonstrate that combining our vocabulary pruning method with LoSparse techniques for encoder layers achieves comprehensive model compression with minimal performance degradation. Experimental results show our approach reduces model size by up to X\% while maintaining Y\% of the original performance across standard benchmarks. This work establishes a practical pathway for creating lightweight, task-specific transformer models without the computational burden of traditional pruning methods.
\end{abstract}

@@

@@

In BERT-style models, the token embedding matrix alone can account for 25--50\% of all parameters, yet vocabulary pruning remains comparatively underexplored. Simple heuristics---such as dropping all tokens not observed in a downstream training corpus---can achieve size reductions, but they fail to capture semantic importance, ignore relationships between pruned and retained tokens, and typically require a retraining step to adapt the model to its slimmed vocabulary. As a result, embedding-level pruning often remains a ``bolt-on'' afterthought rather than an integral, efficient component of model compression.

This paper presents Words That Matter, a fine-tuning-free approach to vocabulary pruning in ModernBERT~\cite{modernbert2023} that directly addresses these gaps. Our method proceeds in two complementary stages. First, we perform a task-specific token importance analysis using TF-IDF scores---balancing how often a token appears in the target dataset against its informativeness across the corpus---to identify and retain only the most critical subword embeddings. Second, rather than relegating all pruned tokens to a generic [UNK] index, we apply semantic clustering over the removed embeddings and map each out-of-vocabulary token to its nearest cluster representative. This preserves nuanced semantic relationships without expanding the model's vocabulary size.

Because our pruning pipeline operates purely post-training and introduces no additional gradient updates, it can be applied to any pre-trained transformer with minimal overhead. Moreover, it dovetails seamlessly with existing encoder-layer pruning techniques---such as LoSparse's~\cite{losparse2023} combined low-rank plus sparse factorization---enabling an end-to-end compression strategy that tackles both embeddings and intermediate representations.

We validate our approach on the GLUE benchmark suite~\cite{wang2018glue}, showing that we can shrink ModernBERT's embedding matrix by up to 20\% while incurring less than 1\% average degradation in downstream accuracy. When combined with LoSparse encoder pruning, overall parameter reduction exceeds 35\% with still under 2\% performance loss---demonstrating that vocabulary and encoder pruning are both necessary and synergistic for building lightweight, high-performance transformer models.

@@

    \item A modular pipeline compatible with any pre-trained encoder model and complementary to structured encoder-layer pruning methods.
\end{itemize}

By placing vocabulary pruning on equal footing with encoder-layer sparsification, Words That Matter opens a practical pathway to more compact, efficient, and deployable transformer models---bringing state-of-the-art NLP within reach of resource-constrained environments.

\section{Related work}
Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements. Most modern LLM compression approaches operate under post-training settings, eliminating the need for resource-intensive retraining. Model compression strategies for LLMs can be categorized into four primary methods:

\begin{enumerate}
    \item Quantization: Reduces the numerical precision of model weights, decreasing memory footprint while maintaining reasonable performance.

@@

    \item Knowledge distillation: Transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size.
\end{enumerate}

These approaches are complementary and can be combined to achieve optimal compression from different perspectives. 
This work focuses on pruning techniques to reduce model size while preserving performance.
For ModernBERT, an encoder-only model, this research examines specific pruning methods for encoder layers and separate techniques for the embedding matrix.

@@

Vocabulary pruning removes rarely used or task-irrelevant tokens to shrink the model's vocabulary (and its embedding matrix). Early work by Abdaoui et al.~\cite{abdaoui2020load} showed that most parameters of multilingual BERT lie in the embedding layer, so trimming unused language tokens can dramatically cut size. In ``Load What You Need,'' they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.

More recently, Nair et al.~\cite{nair2023blade} proposed \textbf{BLADE}: they prune a bilingual mBERT vocabulary to include only tokens from the query and document languages for cross-lingual IR, then apply a small intermediate pretraining step. This pruned model (keeping only needed embeddings) reduced parameters by $\sim$36.5\% versus full mBERT and sped up training by $\sim$30\% and inference by $\sim$55\%, with minimal loss in retrieval quality.

Dorkin et al.~\cite{dorkin2025estonian} also evaluated pruning for an Estonian adaptation: pruning all tokens unused by Estonian led to no degradation on NER, whereas retraining a new tokenizer hurt performance. These studies consistently find that deleting infrequently seen subwords yields large size savings with little accuracy drop.

TextPruner by Shen et al.~\cite{shen2022textpruner} automates vocabulary pruning: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix. This simple ``drop unused tokens'' mode can cut model size and speed up training/inference on that domain.

In sparse-retrieval models like SPLADE, vocabulary pruning is implicit. SPLADE-X~\cite{formal2023spladex} (multi-lingual) restricted output to query-language tokens only, effectively pruning all other vocab. BLADE extends this idea by explicitly building a pruned bilingual model with only the union of query and document subwords, discarding other embeddings entirely.

Across tasks and languages, pruned-vocabulary models often match full models. For example, after pruning, Nair et al.~\cite{nair2023blade} report retrieval effectiveness on par with larger baselines, while achieving much faster indexing. Dorkin et al.~\cite{dorkin2025estonian} explicitly note ``vocabulary pruning has no observable negative effect on the downstream task'' (Estonian NER). These findings suggest vocabulary pruning is an effective way to tailor large models to specific domains or languages.

@@

Pruning the embedding layer often means removing or compressing rows (token vectors) to shrink this heavy component. \textbf{Partial Embedding Matrix Adaptation (PEMA)} by Bousquet et al.~\cite{bousquet2023pema} is one systematic approach: they first note that fine-tuning datasets typically use only a small fraction of the full vocabulary. PEMA therefore builds a partial embedding matrix containing only tokens seen in the task data, training on that smaller matrix. This saves GPU memory during fine-tuning without altering task accuracy. After fine-tuning, the updated embeddings are merged back into the full matrix so the model structure remains unchanged. Experiments show large memory reductions: e.g., BERT$_{\text{BASE}}$'s embedding matrix can shrink by $\sim$47\% on average (and RoBERTa by $\sim$52\%) when only task-relevant tokens are kept. Multilingual models see even more savings (mBERT $\sim$44\% embedding reduction, XLM-R $\sim$64\%) because their vocabularies are huge. Crucially, PEMA reports no drop in downstream performance, making it a practical way to prune embeddings during training.

Another approach is \textbf{embedding compression} via factorization or hashing. Wang et al.~\cite{wang2023lighttoken} propose \textbf{LightToken}, which compresses the entire embedding matrix using low-rank SVD plus a binary residual autoencoder. They note token embeddings are highly redundant (the embedding takes $\sim$21--31\% of BERT/RoBERTa's size). LightToken first applies a rank-$k$ SVD to capture most variation, then learns compact hash codes to reconstruct the residual. In experiments on GLUE, LightToken achieves very high compression (e.g., 25$\times$ smaller embeddings) while preserving accuracy. For BERT$_{\text{BASE}}$, a 25$\times$ embedding compression yields an average GLUE score of 82.9 (baseline 83.0).

These methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance.

Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy. By either dropping unused rows (PEMA/TextPruner) or factoring and quantizing embeddings (LightToken), these methods reduce the $\sim$25--50\% of model parameters in the embedding matrix, enabling lighter models in practice.

@@

Although our focus is vocab/embedding pruning, many complementary pruning methods work at the encoder layers. A classic example is \textbf{attention-head pruning}: Michel et al.~\cite{michel2019sixteen} discovered that a large percentage of attention heads can be removed at test time without significantly impacting performance. Voita et al.~\cite{voita2019analyzing} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.

In practice, toolkit frameworks (like TextPruner) often support head and FFN pruning alongside vocab pruning. Other methods include \textit{movement pruning}~\cite{sanh2020movement} that gradually zeros out small weights during fine-tuning, or \textit{LayerDrop}~\cite{fan2020layerdrop} which stochastically drops whole layers during training.

@@

The proposed hybrid vocabulary pruning method for ModernBERT targets the embedding layer, which constitutes approximately 25\% of the model's parameters. This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that removed tokens can be mapped to semantically similar ones instead of a single unknown token. The method consists of three main components: token importance analysis, selective pruning, and semantic clustering for out-of-vocabulary (OOV) tokens.

\subsection{Token Importance Analysis}
Multiple approaches for token importance estimation were systematically evaluated:

\begin{enumerate}
    \item \textbf{Random Selection}: Tokens are pruned randomly without consideration for importance, serving as a baseline approach.

    \item \textbf{Frequency-Based}: Tokens are ranked by their frequency in the target dataset, with least frequently used tokens pruned first. This approach assumes that rarely used tokens contribute less to model performance.

    \item \textbf{Clustering-Based}: This approach employs agglomerative or k-means clustering on the embedding space to group tokens with similar semantic properties. For each cluster, the token closest to the centroid is retained as a representative, while others are pruned. This preserves semantic diversity across the vocabulary while reducing redundancy.

    \item \textbf{Attention-Based}: This method uses attention patterns from fine-tuning to determine token importance. For each token, attention scores across all layers and heads are aggregated to capture its contextual relevance within the specific task. Tokens receiving consistently higher attention are considered more important for model decisions, providing a more nuanced importance measure that captures both frequency and semantic significance.

    \item \textbf{TF-IDF Based}: Tokens are ranked using Term Frequency-Inverse Document Frequency scores, which balance token occurrence frequency with discriminative power across documents.
\end{enumerate}

The TF-IDF approach demonstrated superior performance and was adopted as the primary method. For this approach, three normalization variants were tested:
\begin{itemize}
    \item Non-normalized TF-IDF (prioritizing rare but task-specific tokens)
    \item L1-normalized TF-IDF (balancing frequency and importance)
    \item L2-normalized TF-IDF (standard approach, providing optimal results)
\end{itemize}

The TF-IDF method effectively identifies task-relevant tokens by assigning higher importance to tokens that frequently appear in specific documents but are less common across the entire corpus. This enables selective vocabulary pruning while preserving tokens critical for maintaining task performance.

\subsection{Hybrid Pruning Method}
The hybrid approach combines importance-based token pruning with semantic clustering for OOV token handling. The method is formalized in Algorithm~\ref{alg:hybrid_pruning}.

\begin{algorithm}[H]
\caption{Hybrid Vocabulary Pruning}
\label{alg:hybrid_pruning}
\begin{algorithmic}[1]
\scriptsize
\Require Model $M$ with vocabulary $V$, dataset $D$, pruning percentage $p$, number of clusters $k$
\Ensure Pruned model $M'$ with reduced vocabulary and OOV mapping

\State // 1. Token Importance Analysis
\State Calculate importance scores for each token in $V$ using TF-IDF on dataset $D$
\State Sort tokens by their importance scores

\State // 2. Pruning Method
\State Keep top $(100-p)\%$ of tokens with highest importance scores
\State Identify tokens to remove: $V_{remove} = V \setminus V_{keep}$

\State // 3. OOV Token Handling
\State Extract embeddings for tokens to be removed: $E_{remove} = M.embeddings[V_{remove}]$
\State Cluster $E_{remove}$ into $k$ clusters using K-means algorithm
\State For each cluster, select token closest to centroid as representative
\State Create OOV mapping: each removed token maps to its cluster representative

\State // 4. Model Modification
\State Construct new vocabulary: $V' = V_{keep} \cup V_{representatives}$
\State Create reduced embedding matrix using embeddings of $V'$
\State Update model $M'$ with reduced embedding layer
\State Modify tokenization process to use OOV mapping for unseen tokens

\State \Return $M'$ with pruned vocabulary and OOV mapping
\end{algorithmic}
\end{algorithm}

\subsection{Out-of-Vocabulary Token Handling}
A key contribution of this approach is the handling of out-of-vocabulary (OOV) tokens. Rather than mapping all pruned tokens to a single UNK token, the semantic properties of the embedding space are utilized to create meaningful representations for OOV tokens.

The clustering process organizes removed tokens into semantic groups, with each group represented by the token closest to the cluster centroid. When the model encounters an OOV token during inference, it maps to the appropriate cluster representative, preserving semantic properties. This approach retains more information than traditional UNK token methods.

\subsection{Implementation Details}
The implementation requires no additional fine-tuning after vocabulary pruning, improving efficiency compared to approaches that require retraining. The method comprises three technical components:

\begin{enumerate}
    \item A token extraction and importance calculation module interfacing with downstream task datasets
    \item An embedding space K-means clustering algorithm creating semantic maps for OOV tokens
    \item A hybrid data collator handling tokenization and OOV mappings during model inference
\end{enumerate}

Four pruning mechanisms are supported:
\begin{itemize}
    \item Frequency-based pruning (retaining most frequent tokens)
    \item Importance-based pruning using non-normalized TF-IDF
    \item Importance-based pruning using L1-normalized TF-IDF 
    \item Importance-based pruning using L2-normalized TF-IDF (default)
\end{itemize}

Experimental results indicate that TF-IDF variants generally outperform pure frequency-based approaches, with L2-normalized TF-IDF providing the optimal balance between task-specific vocabulary retention and general language understanding.

\newpage
\section{Experiments}
% \begin{itemize}
%     \item Multiple pruning percentages (mrpc, mnli, sst2)
%     \item LoSparse + Vocab pruning combined (mrpc, mnli, sst2)
% \end{itemize}
This section evaluates the proposed vocabulary pruning method on the GLUE benchmark, both in isolation and in combination with structured encoder-layer pruning (LoSparse), in order to assess effectiveness across a variety of downstream tasks. The experiments are designed to address the following questions:

\begin{enumerate}
    \item Impact of pure vocabulary pruning: How does performance vary at different pruning ratios?
    \item Comparison of pruning strategies: Does the TF-IDF + OOV hybrid approach outperform simpler baselines (e.g., frequency-based, random selection, clustering, attention-based)?
    \item Interaction with encoder sparsification: What synergistic effects arise when combining embedding-level pruning with encoder-layer sparsification in ModernBERT?
\end{enumerate}

\subsection{Datasets and Metrics}
Eight GLUE development tasks are employed: SST-2, MRPC, MNLI, CoLA, STS-B, QQP, QNLI, and RTE. Classification tasks report accuracy; MRPC additionally reports F1, CoLA reports Matthew correlation coefficient, and STS-B reports Pearson correlation.

\subsection{Baselines and Pruning Methods}
The following configurations are compared:

\begin{itemize}
    \item ModernBERT (Full): No pruning.
    \item Train Tokens Only: Removal of embeddings not observed in fine-tuning data.
    \item Random Selection: Uniform removal of embeddings.
    \item Clustering-Based: Retention of one centroid token per semantic cluster.
    \item Frequency-Based: Pruning of lowest-frequency tokens.
    \item Attention-Based: Removal of tokens with lowest aggregated attention weights.
    \item TF-IDF Based: Pruning according to L2-normalized TF-IDF scores.
    \item Frequency + OOV: Hybrid schemes that remap pruned tokens via clustering representatives.
    \item TF-IDF + OOV: Hybrid schemes that remap pruned tokens via clustering representatives.
    \item LoSparse (20\% encoder pruning): Structured encoder-layer pruning only.
    \item LoSparse + TF-IDF + OOV: Joint embedding and encoder pruning.
\end{itemize}

>>>>

% !TEX root = main.tex
\documentclass[twocolumn]{article}
\usepackage[a4paper,margin=1in]{geometry} 
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{mathptmx}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{arydshln}
\usepackage{float}
\usepackage{cite}
\usepackage{multirow}
\usepackage{graphicx}
\usepackage{url}

@@

\begin{abstract}
Large language models like ModernBERT face deployment challenges due to their computational demands and memory footprint, limiting their use in resource-constrained environments. While various pruning techniques exist, most require resource-intensive fine-tuning, creating a bottleneck for efficient model adaptation. This paper introduces a novel vocabulary pruning approach for ModernBERT that requires no additional fine-tuning, delivering superior efficiency particularly at smaller pruning ratios (5-20\%). The technique specifically targets the embedding layer, which constitutes 25.95\% of the model's parameters, through systematic token importance analysis based on TF-IDF with out-of-vocabulary (OOV) token handling. Experiments on the GLUE benchmark demonstrate that vocabulary pruning alone achieves 20-23\% parameter reduction while maintaining 99.2\% of the original performance, outperforming encoder-based pruning methods at these ratios. 
Benchmark measurements confirm practical benefits: 20\% reduction in storage size, about 15\% reduction in GPU memory, and no impact on inference time. This work establishes an efficient pathway for creating lightweight, task-specific transformer models through offline pre-fine-tuning optimizations rather than computationally expensive retraining.
\end{abstract}

@@

\\ \\

@@

\\ \\
In BERT-style models, the token embedding matrix alone can account for 25.95\% of all parameters in ModernBERT, yet vocabulary pruning remains comparatively underexplored. Simple heuristics---such as dropping all tokens not observed in a downstream training corpus---can achieve size reductions, but they fail to capture semantic importance, ignore relationships between pruned and retained tokens, and typically require a retraining step to adapt the model to its slimmed vocabulary. As a result, embedding-level pruning often remains a ``bolt-on'' afterthought rather than an integral, efficient component of model compression.
\\ \\
This paper presents a fine-tuning-free approach to vocabulary pruning in ModernBERT~\cite{modernbert2023} that directly addresses these gaps. The method proceeds in two complementary stages. First, a task-specific token importance analysis using TF-IDF scores---balancing how often a token appears in the target dataset against its informativeness across the corpus---identifies and retains only the most critical subword embeddings. Second, rather than relegating all pruned tokens to a generic [UNK] index, semantic clustering over the removed embeddings maps each out-of-vocabulary token to its nearest cluster representative. This preserves nuanced semantic relationships without expanding the model's vocabulary size.
\\ \\
Because this pruning pipeline operates purely as a pre-fine-tuning optimization step and introduces no additional gradient updates, it can be applied to any pre-trained transformer with minimal overhead. Moreover, it dovetails seamlessly with existing encoder-layer pruning techniques---such as LoSparse's~\cite{losparse2023} combined low-rank plus sparse factorization---enabling an end-to-end compression strategy that tackles both embeddings and intermediate representations.
\\ \\
Validation on the GLUE benchmark suite~\cite{wang2018glue} shows that ModernBERT's embedding matrix can be reduced by up to 77.34\% (translating to a 20\% total parameter reduction) while incurring less than 1\% average degradation in downstream accuracy. The approach demonstrates superior performance compared to encoder-layer pruning techniques at lower pruning ratios (5-20\%), with practical benefits including 20\% reduction in storage size and 14.83\% reduction in GPU memory requirements. When combined with LoSparse encoder pruning for more aggressive compression, overall parameter reduction exceeds 35\% with still under 2\% performance loss---demonstrating that vocabulary and encoder pruning are both necessary and synergistic for building lightweight, high-performance transformer models.
\\ \\

@@

    \item A modular pipeline compatible with any pre-trained encoder model and complementary to structured encoder-layer pruning methods.
\end{itemize}
By placing vocabulary pruning on equal footing with encoder-layer sparsification, Words That Matter opens a practical pathway to more compact, efficient, and deployable transformer models---bringing state-of-the-art NLP within reach of resource-constrained environments.

\section{Related work}
Model compression techniques enhance the efficiency of large language models (LLMs) by reducing their size and computational requirements. Most modern LLM compression approaches operate under post-training settings, eliminating the need for resource-intensive retraining. Model compression strategies for LLMs can be categorized into four primary methods:
\\ \\
\begin{enumerate}
    \item Quantization: Reduces the numerical precision of model weights, decreasing memory footprint while maintaining reasonable performance.

@@

    \item Knowledge distillation: Transfers knowledge from larger teacher models to smaller student models, enabling competitive performance with reduced architecture size.
\end{enumerate}
These approaches are complementary and can be combined to achieve optimal compression from different perspectives. 
This work focuses on pruning techniques to reduce model size while preserving performance.
For ModernBERT, an encoder-only model, this research examines specific pruning methods for encoder layers and separate techniques for the embedding matrix.

@@

Vocabulary pruning removes rarely used or task-irrelevant tokens to shrink the model's vocabulary (and its embedding matrix). Early work by Abdaoui et al.~\cite{abdaoui2020load} showed that most parameters of multilingual BERT lie in the embedding layer, so trimming unused language tokens can dramatically cut size. In ``Load What You Need,'' they drop languages from mBERT's vocabulary and obtain up to a 45\% reduction in total parameters with comparable accuracy on XNLI.
\\ \\
More recently, Nair et al.~\cite{nair2023blade} proposed \textbf{BLADE}: they prune a bilingual mBERT vocabulary to include only tokens from the query and document languages for cross-lingual IR, then apply a small intermediate pretraining step. This pruned model (keeping only needed embeddings) reduced parameters by $\sim$36.5\% versus full mBERT and sped up training by $\sim$30\% and inference by $\sim$55\%, with minimal loss in retrieval quality.
\\ \\
Dorkin et al.~\cite{dorkin2025estonian} also evaluated pruning for an Estonian adaptation: pruning all tokens unused by Estonian led to no degradation on NER, whereas retraining a new tokenizer hurt performance. These studies consistently find that deleting infrequently seen subwords yields large size savings with little accuracy drop.
\\ \\
TextPruner by Shen et al.~\cite{shen2022textpruner} automates vocabulary pruning: it scans a corpus and removes any tokenizer token not present in the text, deleting its row in the embedding matrix. This simple ``drop unused tokens'' mode can cut model size and speed up training/inference on that domain.
\\ \\
In sparse-retrieval models like SPLADE, vocabulary pruning is implicit. SPLADE-X~\cite{formal2023spladex} (multi-lingual) restricted output to query-language tokens only, effectively pruning all other vocab. BLADE extends this idea by explicitly building a pruned bilingual model with only the union of query and document subwords, discarding other embeddings entirely.
\\ \\
Across tasks and languages, pruned-vocabulary models often match full models. For example, after pruning, Nair et al.~\cite{nair2023blade} report retrieval effectiveness on par with larger baselines, while achieving much faster indexing. Dorkin et al.~\cite{dorkin2025estonian} explicitly note ``vocabulary pruning has no observable negative effect on the downstream task'' (Estonian NER). These findings suggest vocabulary pruning is an effective way to tailor large models to specific domains or languages.

@@

Pruning the embedding layer often means removing or compressing rows (token vectors) to shrink this heavy component. \textbf{Partial Embedding Matrix Adaptation (PEMA)} by Bousquet et al.~\cite{bousquet2023pema} is one systematic approach: they first note that fine-tuning datasets typically use only a small fraction of the full vocabulary. PEMA therefore builds a partial embedding matrix containing only tokens seen in the task data, training on that smaller matrix. This saves GPU memory during fine-tuning without altering task accuracy. After fine-tuning, the updated embeddings are merged back into the full matrix so the model structure remains unchanged. Experiments show large memory reductions: e.g., BERT$_{\text{BASE}}$'s embedding matrix can shrink by $\sim$47\% on average (and RoBERTa by $\sim$52\%) when only task-relevant tokens are kept. Multilingual models see even more savings (mBERT $\sim$44\% embedding reduction, XLM-R $\sim$64\%) because their vocabularies are huge. Crucially, PEMA reports no drop in downstream performance, making it a practical way to prune embeddings during training.
\\ \\
Another approach is \textbf{embedding compression} via factorization or hashing. Wang et al.~\cite{wang2023lighttoken} propose \textbf{LightToken}, which compresses the entire embedding matrix using low-rank SVD plus a binary residual autoencoder. They note token embeddings are highly redundant (the embedding takes $\sim$21--31\% of BERT/RoBERTa's size). LightToken first applies a rank-$k$ SVD to capture most variation, then learns compact hash codes to reconstruct the residual. In experiments on GLUE, LightToken achieves very high compression (e.g., 25$\times$ smaller embeddings) while preserving accuracy. For BERT$_{\text{BASE}}$, a 25$\times$ embedding compression yields an average GLUE score of 82.9 (baseline 83.0).
\\ \\
These methods show that even structured low-rank or quantized pruning of embeddings can yield massive space savings with minimal impact on performance.
\\ \\
Overall, embedding-layer pruning techniques exploit the fact that many token vectors (especially for rare subwords) contribute little to end-task accuracy. By either dropping unused rows (PEMA/TextPruner) or factoring and quantizing embeddings (LightToken), these methods reduce the $\sim$25--50\% of model parameters in the embedding matrix, enabling lighter models in practice.

@@

Although our focus is vocab/embedding pruning, many complementary pruning methods work at the encoder layers. A classic example is \textbf{attention-head pruning}: Michel et al.~\cite{michel2019sixteen} discovered that a large percentage of attention heads can be removed at test time without significantly impacting performance. Voita et al.~\cite{voita2019analyzing} likewise pruned 38 of 48 encoder heads in an NMT transformer (removing $\sim$79\% of heads) with only a 0.15 BLEU loss.
\\ \\
In practice, toolkit frameworks (like TextPruner) often support head and FFN pruning alongside vocab pruning. Other methods include \textit{movement pruning}~\cite{sanh2020movement} that gradually zeros out small weights during fine-tuning, or \textit{LayerDrop}~\cite{fan2020layerdrop} which stochastically drops whole layers during training.

@@

The proposed hybrid vocabulary pruning method for ModernBERT targets the embedding layer, which constitutes approximately 25\% of the model's parameters. This approach is based on the observation that tokens in a vocabulary have varying importance for downstream tasks, and that certain tokens can be selectively removed with minimal impact on performance. The method primarily consists of two main components: token importance analysis and selective pruning, with an optional third component: semantic clustering for out-of-vocabulary (OOV) tokens that can further enhance performance in some cases.
\\ \\
\subsection{Pre-Fine-Tuning Pruning Procedure}
In contrast to methods that require resource-intensive post-pruning fine-tuning, the proposed vocabulary pruning approach operates as a pre-fine-tuning offline optimization step in the model adaptation pipeline. The standard workflow for adapting ModernBERT to a downstream task involves taking the pre-trained base model and fine-tuning it directly on the task data. This research modifies this workflow as follows:
\\ \\
\begin{enumerate}
    \item Task data analysis: The downstream task dataset is processed to extract token statistics.
    \item Token importance calculation: Based on the analysis, tokens are ranked by their importance using one of several methods (frequency, TF-IDF, attention patterns, etc.).
    \item Vocabulary pruning: The embedding layer is pruned by removing less important tokens according to the calculated rankings.
    \item Optional OOV handling: For pruned tokens, semantic clustering may be applied to create mappings to representative tokens.
    \item Model fine-tuning: The pruned model is then fine-tuned on the downstream task using standard procedures.
\end{enumerate}
All pruning techniques follow these common steps:
\begin{enumerate}
    \item Always preserve special tokens ([CLS], [SEP], [PAD], etc.) regardless of their importance scores
    \item Calculate token importance using the technique-specific method
    \item Sort tokens by their importance scores in descending order
    \item Keep the top $(100-p)\%$ highest-scoring tokens, where $p$ is the pruning percentage
    \item Create a reduced embedding matrix using only the retained tokens
\end{enumerate}
This approach front-loads the pruning work to the pre-fine-tuning phase, meaning the actual fine-tuning process remains unchanged and operates on a model that already has a reduced parameter count. The resulting pruned model maintains its standard inference pipeline while benefiting from reduced memory requirements.
\\ \\
The high-level vocabulary pruning procedure, combining importance-based token pruning with optional semantic clustering for OOV token handling, is formalized in Algorithm~\ref{alg:hybrid_pruning}.
\\ \\
\begin{algorithm}[H]
\caption{Vocabulary Pruning Procedure}
\label{alg:hybrid_pruning}
\begin{algorithmic}[1]
\footnotesize
\Require Pre-trained model $M$ with vocabulary $V_{orig}$, task dataset $D$, pruning ratio $p$
\Ensure Pruned model $M'$ with reduced vocabulary

\Function{ExtractTaskVocabulary}{$D, V_{orig}$}
    \State $V_{task} \gets $ unique tokens in dataset $D$
    \State $V_{special} \gets $ special tokens from $V_{orig}$ (e.g., [CLS], [SEP])
    \State \Return $V_{special} \cup V_{task}$ \Comment{Initial vocabulary reduction}
\EndFunction

\Function{RankByImportance}{$V, D$}
    \State Calculate importance scores for tokens in $V$ using selected method
    \State Sort $V$ by descending importance score
    \State \Return sorted token list
\EndFunction

\Function{PruneVocabulary}{$V_{ranked}, p, M$}
    \State $k \gets \lfloor |V_{ranked}| \times (1 - p) \rfloor$ \Comment{Number of tokens to keep}
    \State $V_{keep} \gets $ first $k$ tokens from $V_{ranked}$
    \State $V_{prune} \gets V_{ranked} \setminus V_{keep}$ \Comment{Tokens to be pruned}

    \If{using OOV handling}
        \State Create mapping from pruned tokens $V_{prune}$ to appropriate representatives
        \State \Return $V_{keep}$, mapping
    \Else
        \State \Return $V_{keep}$, null
    \EndIf
\EndFunction

\Function{CreateReducedModel}{$M, V_{keep}, mapping$}
    \State Extract embeddings for tokens in $V_{keep}$ from $M$
    \State Create reduced embedding matrix $E'$
    \State Update model with reduced embedding layer using $E'$
    \If{mapping is not null}
        \State Store mapping for OOV token handling during inference
    \EndIf
    \State \Return updated model
\EndFunction

\State // Main procedure
\State $V_{base} \gets $ \Call{ExtractTaskVocabulary}{$D, V_{orig}$}
\State $V_{ranked} \gets $ \Call{RankByImportance}{$V_{base}, D$}
\State $V_{keep}$, mapping $\gets $ \Call{PruneVocabulary}{$V_{ranked}, p, M$}
\State $M' \gets $ \Call{CreateReducedModel}{$M, V_{keep}$, mapping}

\State \Return $M'$ \Comment{Model ready for fine-tuning}
\end{algorithmic}
\end{algorithm}

\subsection{Token Importance Analysis}
A critical challenge in vocabulary pruning is determining which tokens to retain and which to remove. This research explores and compares several token importance estimation techniques, each with distinct theoretical foundations and practical implications for model performance. These approaches range from simple statistical methods to complex semantic analysis.
\\ \\
\subsubsection{Random Selection}
Tokens are pruned randomly without consideration for importance, serving as a baseline approach. In this method, tokens are selected for removal using uniform random sampling from the full vocabulary.
\\ \\
\subsubsection{Clustering-Based}
This approach employs agglomerative hierarchical clustering on the embedding space to group tokens with similar semantic properties. The algorithm works as follows:
\\ \\
\begin{enumerate}
    \item Each token's embedding vector is normalized to unit length (magnitude of 1), allowing the algorithm to compare tokens based solely on their semantic direction rather than magnitude.
    \item Agglomerative clustering begins with each token as its own cluster, then iteratively merges the closest pairs of clusters.
    \item The merging process uses average linkage, where the distance between clusters is the average distance between all pairs of embeddings across the two clusters.
    \item The algorithm stops when the target number of clusters is reached (determined by the pruning percentage).
    \item For each resulting cluster, the token closest to the centroid is identified and retained as the representative.
    \item All other tokens in the cluster are pruned from the vocabulary.
\end{enumerate}
This method preserves semantic diversity across the vocabulary while reducing redundancy. By selecting representatives closest to cluster centroids, the approach ensures that the preserved tokens maximize coverage of the semantic space. The clustering leverages the property that tokens with embeddings pointing in similar directions (high cosine similarity) tend to have similar semantic meanings, allowing one representative token to effectively substitute for multiple semantically related tokens.
\\ \\
\subsubsection{Frequency-Based}
Tokens are ranked by their frequency in the target dataset, with least frequently used tokens pruned first. This approach operates on the principle that rarely used tokens contribute less to model performance. The technique-specific algorithm steps are:
\\ \\
\begin{enumerate}
    \item Count occurrences of each token in the task-specific dataset
    \item Sort tokens by frequency (most common first)
\end{enumerate}
\subsubsection{Attention-Based}
This method leverages the attention mechanism of transformer models to identify important tokens. The key insight is that tokens receiving higher attention during inference are those the model relies on most heavily when making predictions, providing a direct measure of token importance from the model's perspective. Unlike frequency-based approaches that simply count occurrences, attention-based pruning preserves tokens that meaningfully contribute to the model's decision-making process.
\\ \\
As an illustrative example, the important tokens identified in the SST-2 dataset differ markedly between methods. Frequency-based approaches naturally prioritize common words like articles and prepositions, while attention-based analysis highlights sentiment-laden terms that are semantically relevant to the classification task.
\begin{figure*}[t]
\centering
\begin{minipage}{0.48\textwidth}
  \centering
  \includegraphics[width=\textwidth]{images/wordcloud_frequency.png}
\end{minipage}%
\hfill
\begin{minipage}{0.48\textwidth}
  \centering
  \includegraphics[width=\textwidth]{images/wordcloud_attention.png}
\end{minipage}
\caption{Comparison of top-ranked tokens for SST-2 task using frequency-based (left) versus attention-based (right) importance metrics.}
\label{fig:wordcloud-comparison}
\end{figure*}
The visualization in Figure \ref{fig:wordcloud-comparison} demonstrates how attention-based pruning better captures task-relevant information, preserving tokens that carry greater semantic weight for the specific downstream task.
\\ \\
The technique-specific algorithm steps are:
\begin{enumerate}
    \item First fine-tune the base model on the target task to learn task-specific attention patterns
    \item Process the task dataset through this fine-tuned model, capturing attention matrices from all layers and heads
    \item For each token, aggregate the attention it receives across all contexts where it appears
    \item Normalize attention scores by token frequency to avoid bias toward common tokens
\end{enumerate}
\subsubsection{TF-IDF Based}
Tokens are ranked using Term Frequency-Inverse Document Frequency (TF-IDF) scores, which balance token occurrence frequency with discriminative power across documents. This method prioritizes tokens that appear frequently in specific documents but are rare across the corpus—capturing task-specific terminology while filtering out ubiquitous tokens that carry less semantic value.
\\ \\
The TF-IDF approach is particularly effective for tasks with distinct document categories (like sentiment classification or topic detection) where certain tokens strongly differentiate between classes. Unlike simple frequency counts, TF-IDF prevents common but semantically shallow tokens (articles, conjunctions, etc.) from dominating the pruned vocabulary.
\\ \\
The technique-specific algorithm steps are:
\begin{enumerate}
    \item Tokenize all examples in the task dataset
    \item Calculate term frequency (TF) for each token in each document:
        \begin{equation}
        \text{TF}(t,d) = \frac{\text{count of term $t$ in document $d$}}{\text{total number of terms in document $d$}}
        \end{equation}
    \item Calculate inverse document frequency (IDF) across the corpus:
        \begin{equation}
